---
title: "Introduction to R"
author: Paul Hewson
date: August 5th 2014
output: ioslides_presentation
---

# Introduction to R


## What is R?

- An environment for statistical data analysis
- Interactive (command line) data analysis
- Programming (scripting?) language

## History of R

- Ross Ihaka and Robert Gentleman in Aukland wanted a teaching alternative to S (S-Plus).   Worked under GPL and allowed a core team to form.
- There are variants to enable faster/less memory restricted - for example Luke Tierney has been working on a byte code compiler


## How do you use R?

- Emacs has a great IDE - [Emacs speaks statistics](http://ess.r-project.org/)
- [Rstudio](www.rstudio.com) is a very comprehensive IDE
- Various GUI builders possible - if you wanted to construct button clicking possibilities for non-programming colleagues e.g. Shiny, Tcl-Tk (and many many more)

## How do you develop in R

- There are > 3000 add on packages, and derivative projects such as [Bioconductor](http://bioconductor.org/)
- There are two methods (S3 and S4) formalising some kind of object oriented programming
- There are many ways to drop down into Fortran or C/C++ if you need to speed up a particular routine - see [writing R extensions](http://cran.r-project.org/doc/manuals/R-exts.pdf)
- There are many add on routines for parallel functioning (see [High performance and parallel R task view](http://cran.r-project.org/web/views/HighPerformanceComputing.html)

## What are we going to do this afternoon

- Introduce R to do some statistical modelling
- At the end, we will have a quick plenary on the statistical concepts
- For now, this is a programming / command line scripting exercise.




# Basic R


## Assignment to variables
```{r}
x <- 1
y <- 1
x + y
z <- x + y
y <- 2
z
```
## Vectors and arrays


```{r}
x <- c(1,3)
y <- matrix(c(1,2,3,4),2,2)

x * y
x %*% y
```


## Looping

- This should be familiar, but don't.

```{r}
out <- rep(0,10)
#out <- vector("numeric", 10)
for (i in 1:10){
out[i] <- i^2
}
out
sum(out)
```

## Apply


```{r}
x <- c(1:10)
xsquare <- 0

output <- vapply(x,function(x){x^2}, xsquare)
output
sum(output)
```


## Now do that again

- With a sensible problem
- With a large number of $n$
- Compare timings



# Data

## Reading some data into R

```{r}
#nyc.df <- read.csv("http://dl.dropboxusercontent.com/u/12691674/stat2402/nyc.csv")

library(RCurl)
data <- getURL("http://dl.dropboxusercontent.com/u/12691674/stat2402/nyc.csv", ssl.verifypeer=0L, followlocation=1L)
writeLines(data,'tempnyc.csv')
nyc.df <- read.csv('tempnyc.csv', row.names = 1)
head(nyc.df)
```


## Everything is an object


```{r}
summary(nyc.df)
```

## Now fit a model

```{r}
nyc.lm1 <- lm(Price ~ Food + Decor + Service, data = nyc.df)
summary(nyc.lm1)
```

## Doing data analysis in R

- Various approaches
- Keep the script file and data set - the analysis is reproducible
- Many tools (Sweave, knitr) make literature programming possible
- Good books advise on approaches (see manuals which come with R as a starting point)



# Object oriented programming


## Estimating $\pi$ by Monte Carlo methods

```{r}
mcpi.default <- function(n,a=-1,b=1){
  x <- runif(n,a,b)
  y <- runif(n,a,b)
  inside <- length(x[x^2+y^2 < 1])
  result <- list(x=x,y=y,pi.est=inside/n*4)
  class(result) <- "mcpi"
  result
}  
  
mcpi <- function(n,a,b){UseMethod("mcpi")}
```


## We now just needs methods for this object

```{r}
print.mcpi <- function(out){
  print(out$pi.est)
}

plot.mcpi <- function(out){
  mycol <- rep("red", length(out$x))
mycol[out$x^2 + out$y^2 < 1] <- "blue"
  plot(out$x, out$y, pch = 16, col = mycol, 
       xlab = "x", ylab="y", main= "Monte Carlo simulations")
  }
```

## And useage is simple

```{r}
a.simulation <- mcpi(1000)
print(a.simulation)
plot(a.simulation)
str(a.simulation)
```

# What now

## There are three tutorial sheets to try out

- Rdata: introduces some data handling / model fitting
- Rprogramming: sets a programming exercise
- SomeGGPlotIdeas: introduces another way of visualising data

[https://github.com/phewson/rintro](https://github.com/phewson/rintro)


Pick a sheet and start working!






