---
title: "Introduction to R"
author: Paul Hewson
date: August 5th 2014
output: ioslides_presentation
---

# Introduction to R


## What is R?

- An environment for statistical data analysis
- Interactive (command line) data analysis
- Programming (scripting?) language

## History of R

- Ross Ihaka and Robert Gentleman in Aukland wanted a teaching alternative to S (S-Plus).   Worked under GPL and allowed a core team to form.
- There are variants to enable faster/less memory restricted - for example Luke Tierney has been working on a byte code compiler


## How do you use R?

- Emacs has a great IDE - Emacs speaks statistics
- Rstudio is a very comprehensive IDE
- Various GUI builders possible - if you wanted to construct button clicking possibilities for non-programming colleagues e.g. Shiny, Tcl-Tk (and many many more)

## How do you develop in R

- There are > 3000 add on packages, and derivative projects such as Bioconductor
- There are two methods S3 and S4 formalising some kind of object oriented programming
- There are many ways to drop down into Fortran or C/C++ if you need to speed up a particular routine
- There are many add on routines for parallel functioning (see relevant task view)

## What are we going to do this afternoon

- Introduce R to do some statistical modelling
- At the end, we will have a quick plenary on the statistical concepts
- For now, this is a programming / command line scripting exercise.




# Basic R


## Assignment to variables
```{r}
x <- 1
y <- 1
x + y
z <- x + y
y <- 2
z
```
## Vectors and arrays


```{r}
x <- c(1,3)
y <- matrix(c(1,2,3,4),2,2)

x * y
x %*% y
```


## Looping

- THis should be familiar, but don't.





```{r}
out <- rep(0,10)
#out <- vector("numeric", 10)
for (i in 1:10){
out[i] <- i^2
}
out
sum(out)
```

```{r}
x <- c(1:10)
xsquare <- 0

output <- vapply(x,function(x){x^2}, xsquare)
output
sum(output)
```


## Now do that again

- With a sensible problem
- With a large number of $n$
- Compare timings



# Data

```{r}
#nyc.df <- read.csv("http://dl.dropboxusercontent.com/u/12691674/stat2402/nyc.csv")
#write.csv(nyc.df, row.names = FALSE, "c:\\Users\\phewson\\nyc.csv")
nyc.df <- read.csv("c:\\Users\\phewson\\nyc.csv")

head(nyc.df)

pairs(nyc.df[,-c(1,2,7)])
nyc.lm1 <- lm(Price ~ Food + Decor + Service, data = nyc.df)
summary(nyc.df)
plot(nyc.lm1)
```








