The Beverton-Holt model
=======================


We are going to walk through an example of a non-linear model, specifically the so-called ``Beverton-Holt'' model.   This is a model for fish recruitment.   We are interested in the Spawning stock (spawning biomass) and the resulting number of fish (recruitment)

\begin{equation}
R = f(S, (\alpha, k)) = \frac{\alpha S}{1 + S/k}
\end{equation}


The data we are going to consider are in the R library \texttt{nlrwr}.   

```{r, warning=FALSE, echo=FALSE}
#install.packages("nlrwr")
library(nlrwr)
data(M.merluccius)
```


Because the data are contained within the package, we need to use data() to make the data visible.


Plotting the data
-----------------


### Exercise

Start by plotting the data.   

- The standard R function plot() will work find.

- You want to plot num.fish on the y axis and spawn.biomass on the x axis.

- You can set the axis labels and title with xlab = "label", ylab = "label" and main = "Title" within the function call



### Solution


```{r}
plot(num.fish ~ spawn.biomass, data = M.merluccius, 
     xlab = "Spawning biomass (1000 tonnes)", 
     ylab = "Recruitment (million fish)")
```



Fitting the model
-----------------


For non-linear models, you need to specify the model, and use a function called nls.   Any problems type ?nls for help.


```{r}
try(mm.bh <-  nls(num.fish ~ spawn.biomass * alpha/(1+spawn.biomass/k), 
              data = M.merluccius, 
              start = list(alpha = 0, k = 0)))

```


Note for non-linear models you need to specify initial values of the parameters.   The example given here does not have suitable values - can you find suitable ones.   In order to do this it helps to know what the parameters mean.   We consider the proliferation rate and carrying capacity of the population.



### Solution


```{r}
mm.bh <-  nls(num.fish ~ spawn.biomass * alpha/(1+spawn.biomass/k), 
              data = M.merluccius, 
              start = list(alpha = 5, k = 50))

```




There is a rather nice package called *nlstools* which has a convenient plotting function that lets us plot the data AND the model we just fitted.


```{r}
require(nlstools)
overview(mm.bh)
plotfit(mm.bh)
```

So far so good.   But now we want to do ``inference'' on those parameters.


The first thing we can do (for ``pedagogic purposes'') is to plot a contour of RSS values for possible combinations of the two parameters.

```{r}
contour <- nlsContourRSS(mm.bh)
plot(contour)
```

This is intended more to give us some idea that the function we have been maximising has rather a complicated relationship with the two parameter values.   And hence that we might need to be careful with the interval estimates.


Interval Estimates
-------------------

We've talked about confidence intervals.   There are different ways of computing these.

### Wald Intervals

If you are very lucky, you might have the *nlrwr* package available to you - if so it calculates the Wald intervals for you.   If not you have to do this by hand (which is better exam practice anyway).

```{r}
require(nlrwr) ## lucky if this works
confint2(mm.bh)

summary(mm.bh)
```


Hopefully though, that contour plot warned you that these estimates might be problematic.

### Profile intervals

We met these profile intervals when we examined non-linear models.   So this should be straightforward, except that now you know exactly where the lines come from.

The profile t-statistic is obtained by calculating:

$$\tau(\beta_j) = \mbox{sign} (\beta_j - \hat{\beta_j}) \frac{\sqrt{RSS(\beta_j) - RSS(\hat{\beta_j})}}{s}$$

So we obtain the lines from relevant quantiles of a t-distribution with suitable degrees of freedom.

```{r}
require(nlstools)
mm.bh.profile <- profile(mm.bh)
plot(mm.bh.profile)

confint(mm.bh)
```




### Bootstrapping


For the method we are going to use (pairs based resampling, percentile based confidence intervals), bootstrapping a non-linear model is trivial.


```{r}
require(nlstools)
mm.bh.boot <- nlsBoot(mm.bh)
summary(mm.bh)
summary(mm.bh.boot)

plot(mm.bh.boot)

```

Extension Activity
------------------

### Reparameterisation

This is where life get's interesting.   If you search the web you can find alternative formulations for the Beverton-Holt model.   Can you find and fit a formulation that has better behaving Wald intervals.

### Alternatives

- Deriso model:

$$(S, (\alpha, c, k)) = \alpha S \left( 1 -\frac{cS}{k} \right)^{1/c}$$

(Hint, try start =  list(alpha = 4.4, k = 106, c=0.86))

- Ricker model:

$$f(S, (\alpha, k)) = \alpha S \exp(-kS)$$
(Hint: try start= list(alpha = 5, k = 50))

- Shepherd model

$$f(S, (\alpha, c, k)) = \frac{\alpha S}{1 + \left( \frac{S}{k} \right)^c}$$

(Hint: try start = list(alpha = 3.87, k=61.72, c = 2.25), also add within the function call \texttt{control=nls.control(maxiter=100)}
